---
title: "Parameterize an Albers Equal Area projection"
author: "Michael France Nelson"
date: "May 11, 2020"
output: html_document
---

```{r setup, include=FALSE}
require(ggplot2)
require(sf)
require(sp)
# require(raster)
```


# This page describes how to choose appropriate parameters for an Albers Equal Area conic projection for a region of interest.


# Scenario

- You want to choose an appropriate Albers Equal Area projection for some spatial data.
- You need a practical way to choose parameter values use for the projection, based on the locaiton of your data.


# Summary

Use the following steps to find the four parameters you need to define an acceptable Albers Equal Area projection for your data:

<ol>
<li> Get the lat/long coordinates of the bounding box of your data </li>
<li> Calculate:</li><ol>
<li> The longitude at the midpoint of the bounding box </li>
<li> The latitude at the midpoint of the bounding box </li>
<li> The min and max latitudes of the bounding box </li></ol>
</ol>

The four parameters you need are:

- `lat_1` = min latitude of the bounding box + 1/6 latitude range
- `lat_2` = max latitude of the bounding box - 1/6 latitude range
- `lat_0` = the latitude at the midpoint
- `lon_0` = the longitude at the midpoint

Assemble the proj4 string using your parameter values:

“+proj=aea +lat_1=_____ +lat_2=_____ +lat_0=_____ +lon_0=_____”




# Background {.tabset}

## proj4

To see the proj4 string format, here's an example (the AEA projection used by USGS)

"+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"

This example has values for all of the parameters, optional and required.




## Required parameters {.tabset}

First standard parallel.

- Format: +lat_1=<value>
- Defaults to 0.0.

Second standard parallel.

- Format: +lat_2=<value>
- Defaults to 0.0.




## Optional parameters  {.tabset}

### Longitude of projection center.

- Format: +lon_0=<value>
- Defaults to 0.0.

### Latitude of projection center.

- Format: +lat_0=<value>
- Defaults to 0.0.

### Ellipsoid

- Format: +ellps=<value>
- Defaults to “GRS80”.

### Radius of the sphere

Given in meters.

If used in conjunction with +ellps +R takes precedence.

- Format +R=<value>

### False easting.

- Format: +x_0=<value>
- Defaults to 0.0.

### False northing.

- Format: +y_0=<value>
- Defaults to 0.0.




## Some references

https://proj.org/operations/projections/aea.html
https://www.bluemarblegeo.com/knowledgebase/GeoCalcPBW/Content/ClassDef/Projection/Projections/AlbersEqualArea.html
http://www.georeference.org/doc/albers_conical_equal_area.htm




# Data Setup

We'll use polygon data for the state borders of Massachusetts as an example data set.

Package `spData` has an `sf` object containing polygon data for all the US State borders.

```{r}
require(spData)
mass_border_sf = subset(us_states, NAME == "Massachusetts")
ggplot(mass_border_sf) + geom_sf()
```

We'll make a `SpatialPolygonsDataFrame` object to demo the procedure in both the `spatial` and `sf` paradigms
```{r}
mass_border_sp = as(mass_border_sf, "Spatial")
plot(mass_border_sp)
```

We can check the current projection:
```{r}
st_crs(mass_border_sf)
proj4string(mass_border_sp)
```






# Procedure {.tabset}


## Determine the lon/lat values of the bounding box of your area of interest

```{r}
bbox(mass_border_sp)
st_bbox(mass_border_sf)
```

## The "one-sixth" rule {.tabset}

A rule of thumb is to choose standard parallels at 1/6 and 5/6 the range of latitudes.

These standard parallels are the required parameters to specify an Albers Equal Area projection:

### In `sp`

`bbox()` returns the bounding box coordinates in a table.
We can access the `ymin` and `ymax` slots using array indices:

`extent()` returns an **S4** `Extent` object.  We can access the `ymin` and `ymax` slots using the `@` symbol:

```{r one_sixth_rule_sp}

bbox(mass_border_sp)[1, ]

# diff(bbox(mass_border_sp)[2, ])
# yrange_sp = ymax_sp - ymin_sp

bbox_mass_sp = bbox(mass_border_sp)

ymin_sp = bbox_mass_sp[2, 1]
ymax_sp = bbox_mass_sp[2, 2]
yrange_sp = ymax_sp - ymin_sp

# yrange_sp = diff(bbox(mass_border_sp)[2, ])

proj4_aea_sp =
  paste0(
    "+proj=aea",
    " +lat_1=",
    ymin_sp + yrange_sp / 6,
    " +lat_2=",
    ymax_sp - yrange_sp / 6)
```



### In `sf`

The `st` bbox implementation does not return a `S4` object, but we can get the y range using the dollar sign:

```{r one_sixth_rule_sf}
ymin_sf = st_bbox(mass_border_sf)$ymin
ymax_sf = st_bbox(mass_border_sf)$ymax

yrange_sf = ymax_sf - ymin_sf

proj4_aea_sf =
  paste0(
    "+proj=aea",
    " +lat_1=",
    ymin_sf + yrange_sf / 6,
    " +lat_2=",
    ymax_sf - yrange_sf / 6)
```



### Plot

Plot to check if it worked:

```{r}
plot(spTransform(mass_border_sp, proj4_aea_sp))
ggplot(st_transform(mass_border_sf, proj4_aea_sp)) + geom_sf()
```

Something's wrong!


## Specify projection center

We didn't specify the projection center longitude or latitude.  Both default to 0, which is nowhere near Massachusetts!


For simplicity, use the long/lat coords in the center of the bounding box.

In `sp`:

```{r}
mean(bbox_mass_sp[1, ])
mean(bbox_mass_sp[2, ])
```

In `sf`:

```{r}
bbox_mass = st_bbox(mass_border_sp)
mean(bbox_mass$xmin, bbox_mass$xmax)
mean(bbox_mass$ymin, bbox_mass$ymax)
```


Append the parameters to our existing projections:

```{r}
proj4_aea_sp_2 = paste0(
  proj4_aea_sp,
  " +lat_0=", 
  mean(bbox_mass_sp[2, ]),
  " +lon_0=", 
  mean(bbox_mass_sp[1, ]))

proj4_aea_sf_2 = paste0(
  proj4_aea_sf,
  " +lat_0=", 
  mean(bbox_mass$ymin, bbox_mass$ymax),
  " +lon_0=", 
  mean(bbox_mass$xmin, bbox_mass$xmax))
```



## Plot Again

```{r}
plot(spTransform(mass_border_sp, proj4_aea_sp_2))
ggplot(st_transform(mass_border_sf, proj4_aea_sf_2)) + geom_sf()
```

Success!



















